Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference

### 2017-06-28 workflow/workflow.Rmd
Edits to text.

---
title: "Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, UC Berkeley
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
  - name: Kelly Street
    affilitation: Graduate Group in Biostatistics, UC Berkeley
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, UC Berkeley
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics, UC Berkeley, and Department of Statistics, UC Berkeley
abstract: Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized.
keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, clustering, lineage inference, statistical models, workflow, R, Bioconductor
bibliography: ref.bib
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
---

@@

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression; and (4) DE analysis within and between the lineages. Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examineCombineMany)), where columns correspond to cells and rows to different clusterings. Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters across different clusterings that show large overlap. The first row correspond to a consensus clustering across all candidate clusterings.

@@

We now want to demonstrate how to use the R software package `slingshot` to infer branching lineages and order cells by developmental progression. We connect the clusters identified by `RSEC` with a minimum spanning tree (MST) to learn the global lineage structure. Then, we refine this structure and order cells using highly stable simultaneous principal curves to infer smooth, branching lineages.

From the original published work, we know that the start clusters should be the clusters with HBCs and end clusters the clusters with MV, mOSN, and mSUS cells. Additionally, we know that the clusters with the GBCs should be a junction before the differentiation between MV and mOSN cells. See Figure \@ref(fig:stemcelldiff). The correspondance between the clusters we found and the original clusters is as follow

@@

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBCs | blue | original 1, 5
c2 | GBCs | green | original 2, 3, 11
c3 | mSUS | red | original 4, 7
c4 | mOSN | orange | original 9, 12
c5 | Immature Neuron | purple | original 10, 14
c6 | MV | brown | original 15

To order the clusters and find lineages, we use package `slingshot` where the input is the MDS of low dimensional matrix `W` where the number of dimensions used for MDS is `k = 4`. We found that slingshot algorithm is stable to the number of dimensions `k` of the MDS. Here, we use the unsupervised version of slingshot where we only provide the information of the start cluster but not the end clusters. Using slingshot, we recovered the order of the clusters found in the published work. See Figures \@ref(fig:curves) and \@ref(fig:tree).

The two steps of the slingshot algorithm are implemented in the functions 'getLineages' and 'getCurves'. The first takes the low-dimensional representation of the cells and the vector of cluster labels identified above. It fits an MST to the clusters and identifies lineages as paths through this tree. The output of 'getLineages' is a 'SlingshotDataSet' object containing all the information used to fit the tree and identify lineages. 'getCurves' takes this object as input and fits simultaneous principal curves to the identified lineages. These functions can be run separately, as below, or jointly by the wrapper function, 'slingshot'.

@@

In the workflow, we recover a reasonable pseudotime ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. 'getLineages' has the option for the user to provide known end cluster(s). Here is the code to use slingshot in a supervised setting:

@@

After ordering the cells, we are interested in finding genes that have non-constant expression patterns over pseudotime. We can see global trends in expression by subsetting our original data to the most variable genes and plotting their expression values in a heatmap, ordered by pseudotime.

More formally, we are interested in testing for variable expression over pseudotime. A general additive model (GAM) is a flexible, nonlinear way to test the relationship between a gene's normalized expression values and pseudotime. We implemented this model for each gene and show the top 100 genes by p-value in the heatmap below.

@@

This workflow provides a tutorial for the downstream analysis of scRNA-seq data in R. The workflow covers four main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression; and (4) DE analysis within and between the lineages. The workflow is general and flexible, allowing the user to sustitute the statistical method used in each step by a different method. We hope our proposed workflow will ease the technical aspect of scRNA-seq data analysis and help with the discovery of novel biological insights.

@@

# Author contributions
In order to give appropriate credit to each author of an article, the individual contributions of each author to the manuscript should be detailed in this section. We recommend using author initials and then stating briefly how they contributed.

# Competing interests
All financial, personal, or professional competing interests for any of the authors that could be construed to unduly influence the content of the article must be disclosed and will be displayed alongside the article. If there are no relevant competing interests to declare, please add the following: 'No competing interests were disclosed'.

# Grant information
Please state who funded the work discussed in this article, whether it is your employer, a grant funder etc. Please do not list funding that you have that is not relevant to this specific piece of research. For each funder, please state the funder’s name, the grant number where applicable, and the individual to whom the grant was assigned. If your work was not funded by any grants, please include the line: 'The author(s) declared that no grants were involved in supporting this work.'

>>>>

---
title: "Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, UC Berkeley
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
  - name: Kelly N. Street
    affilitation: Graduate Group in Biostatistics, UC Berkeley
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, UC Berkeley
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics, UC Berkeley, and Department of Statistics, UC Berkeley
abstract: Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized.
keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, clustering, lineage inference, statistical models, workflow, R, Bioconductor
bibliography: ref.bib
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
---

@@

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis within and between the lineages. Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examineCombineMany)), where columns correspond to cells and rows to different clusterings. Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters that show large overlap accross rows. The first row correspond to a consensus clustering across all candidate clusterings.

@@

We now demonstrate how to use the R software package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage. The method comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves. The approach in (1) allows us to identify any number of novel lineages while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.

The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`. The first takes a low-dimensional representation of the cells and a vector of cluster labels. It fits an MST to the clusters and identifies lineages as paths through this tree. The output of `getLineages` is an object of class `SlingshotDataSet` containing all the information used to fit the tree and identify lineages. The function `getCurves` then takes this object as input and fits simultaneous principal curves to the identified lineages. These functions can be run separately, as below, or jointly by the wrapper function, `slingshot`.

From the original published work, we know that the start cluster should correspond to HBCs and the end clusters to MV, mOSN, and mSUS cells. Additionally, we know that GBCs should be at a junction before the differentiation between MV and mOSN cells (Figure \@ref(fig:stemcelldiff)). The correspondance between the clusters we found here and the original clusters is as follows.

@@

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | blue | original 1, 5
c2 | GBC | green | original 2, 3, 11
c3 | mSUS | red | original 4, 7
c4 | mOSN | orange | original 9, 12
c5 | Immature Neuron | purple | original 10, 14
c6 | MV | brown | original 15

To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`. We found that the slingshot results were robust to the number of dimensions `k` for the MDS (we tried `k=` [SD: Add values of `k` that were considered.]). Here, we use the unsupervised version of slingshot, where we only provide the identity of the start cluster but not of the end clusters. This implementation recovers the lineages found in the published work (Figures \@ref(fig:curves) and \@ref(fig:tree)).

@@

In the workflow, we recover a reasonable pseudotime ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `getLineages` has the option for the user to provide known end cluster(s). Here is the code to use `slingsho` in a supervised setting.

@@

After assigning the cells to lineages and ordering them within lineages, we are interested in finding genes that have non-constant expression patterns over pseudotime. We can see global trends in expression by subsetting our original dataset to the most variable genes and displaying their expression values in a heatmap, with cells ordered by pseudotime.

More formally, we are interested in testing for variable expression over pseudotime. A general additive model (GAM) is a flexible, nonlinear way to model the relationship between a gene's normalized expression measures and pseudotime. We implemented this model for each gene and show the top 100 genes by p-value in the heatmap below.

@@

This workflow provides a tutorial for the downstream analysis of scRNA-seq data in R. The workflow covers four main steps: (1) dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates; (2) robust and stable clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis within and between the lineages. The workflow is general and flexible, allowing the user to sustitute the statistical method used in each step by a different method. We hope our proposed workflow will ease the technical aspect of scRNA-seq data analysis and help with the discovery of novel biological insights.

@@

# Author contributions

[SD: Please check.]

FP, DR, EP, and KNS performed the data analaysis. DR, SD, and EP supervised the research. All authors contributed to the writing of the manuscript.

# Competing interests

No competing interests were disclosed.

# Grant information

[SD: Please check.]

DR, KNS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (grant U01 MH105979, PI: John Ngai). KNS was supported by a training grant from the National Human Genome Research Institute (T32000047).
